---
title: CoDA workflow for evaluating the nutrient compositions of cereal grains
author: M.G. Walsh, R. Rodríguez Iglesias, P.C. Nalivata and M.R. Broadley
date: "Last compiled on `r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    css: style.css
---

```{r, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction

Most models of species distributions, photosynthesis, nutrient uptake, yield, trophic dynamics, crop management and/or dietary requirements assume no interaction between individual mineral nutrient elements and typically describe the associated processes as a function of only the potentially most limiting ones (i.e., the [Blackman limitation](https://www.biologydiscussion.com/photosynthesis/blackmans-law-of-limiting-factors-and-his-criticism-photosynthesis/23006)). However, the Blackman limitation does not appear to hold generally. In agricultural systems, the overall nutrient composition is likely to affect both crop yields as well as the overall nutritional quality of plant products that are subsequently consumed by animals, including humans.

Compositional Data Analysis ([CoDA](http://www.compositionaldata.com/)) refers to the analyses of compositional data (CoDa), which are defined as vectors with strictly positive components whose sum is constant (e.g., 1 or 100%). The terms also cover all other *parts* of a whole (so-called *subcompositions*), which carry only relative information, including parts per concentration unit (e.g., ppm or mcg), but potentially also their molar equivalents. Arguably, all multivariate concentration data should be treated CoDa because as a whole they must sum to the per unit total of all the constituent parts e.g., in the case of ppm, to 10^6^ mg kg^-1^. The safe assumption with these types of data is that the parts of interest within the unit total cannot be expected to behave completely independently of all the other constituent parts. 

The original approach to CoDA developed by [J. Aitchison (1982)](https://rss.onlinelibrary.wiley.com/doi/10.1111/j.2517-6161.1982.tb01195.x) uses logarithmic ratios (LRs) of the compositional parts (mineral nutrient elements in this case) as the fundamental starting point for CoDa description and modeling. There is also a paper by [M. Greenacre (2021)](https://www.annualreviews.org/doi/pdf/10.1146/annurev-statistics-042720-124436) openly available, which provides excellent guidance, as well as [R scripts](https://github.com/michaelgreenacre/CODAinPractice), about the practical applications and interpretation of CoDa. We shall be following and testing Greenacre's pragmatic advice and approach here on a relatively large, cereal, mineral nutrient dataset. This notebook is maintained on [Github](https://github.com/mgwalsh/Bioavailability/blob/master/CoDa_workflow.Rmd), and you can fork and alter it from there as you see fit.

# General data setup

The data we will be using were generated by the [GeoNutrition project](http://www.geonutrition.com) in Malawi. They consist of 1,808 cereal grain samples that were collected for maize (1,606), pearl millet (31), rice (54), and sorghum (117) in in proportion of their geographical occurrence. The nationally representative cropland sampling frame and the laboratory methods that were used in the data collections are described in ([Gashu et al., (2021)](https://doi.org/10.1038/s41586-021-03559-3)). 

The data include [ICP-MS](https://en.wikipedia.org/wiki/Inductively_coupled_plasma_mass_spectrometry) concentration measurements  of the respective cereal mineral nutrient contents (in ppm), including: Na, Mg, P, S, K, Ca, Cr, Mn, Fe, Co, Cu, Zn, Se, Mo, and I. They also include the georeference of where the samples were collected (lon/lat, EPSG:3857), the dominant cereal crop at that location, and if, or not, fertilizer and/or lime were used in growing the crop.

## Data download

To actually run this notebook, you will need to load the packages indicated in the chunk directly below to download and assemble the data. The chunk (not shown but in the markdown document) also generates an overview map of where the samples were collected in Malawi ... you can click and zoom into the individual sampling locations.

```{r, results = 'hide'}
# Package names
packages <- c("osfr", "dplyr", "leaflet", "compositions", "vegan", "easyCODA", "archetypes")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))

# Create a data folder in your current working directory
dir.create("MW_coda", showWarnings=F)
setwd("./MW_coda")
dir.create("Results")

# Data download from OSF at: https://osf.io/btxyc/
osf_retrieve_file("btxyc") %>% osf_download() ## only do this once, else erase ./MW_coda before running again
codat <- read.table("MW_grain_coda.csv", header=T, sep=",")
```

```{r, echo = FALSE}
# Grain sample locations
w <- leaflet() %>%
  setView(lng = mean(codat$lon), lat = mean(codat$lat), zoom = 7) %>%
  addProviderTiles(providers$OpenStreetMap.Mapnik) %>%
  addCircleMarkers(codat$lon, codat$lat, clusterOptions = markerClusterOptions())
w ## plot widget 
```

## CoDa conversions

This section of the notebook sets up the CoDa data frame that includes 2 amalgamations: mineral **macro-nutrients** (i.e., Na, Mg, P, S, K, and Ca ... in atomic number order), and **micro-nutrients** (i.e., Cr, Mn, Fe, Co, Cu, Zn, Se, Mo, and I). Note that the difference between *macro* and *micro* here is only intended to reflect the recommended daily dietary values (DDV) that should be consumed by people, and not their overall nutritional importance and/or benefits. The compositional macro / micro differences may also prove to be useful for spatial mapping and/or soil property and MIR spectral predictions of nutrient compositions at a later stage. Those subsequent steps will be described in separate notebooks.

This next chunk [closes](https://econ-papers.upf.edu/papers/1554.pdf) the CoDa, and calculates *centered logratios* ([CLR](https://www.rdocumentation.org/packages/compositions/versions/2.0-1/topics/clr)), which can then be analyzed, summarized and interpreted with conventional univariate, multivariate, and data modeling methods ([Greenacre, 2021](https://www.annualreviews.org/doi/pdf/10.1146/annurev-statistics-042720-124436)). Note that you could also use other logratios such as the [ALR](https://www.rdocumentation.org/packages/compositions/versions/2.0-1/topics/alr), or [ILR](https://www.rdocumentation.org/packages/compositions/versions/2.0-1/topics/ilr) and/or different amalgamations here, but we shall leave those possibilities for you to explore. We also write out the resulting data frame to your `./MW_coda/Results` directory should you wish to process it in software other than R.

```{r}
# Calculate the CLRs
varc <- c("Na", "Mg", "P", "S", "K", "Ca", "Cr", "Mn", "Fe", "Co", "Cu", "Zn", "Se", "Mo", "I")
combo <- codat[varc]
combo <- combo / rowSums(combo) ## close composition
combo <- as.data.frame(clr(combo)) ## centered log ratio (clr) transform

# Define the (macro / micro) amalgamation and add it to the overall nutrient composition
varm <- c("Na", "Mg", "P", "S", "K", "Ca")
macro <- rowSums(combo[varm])
combo <- cbind(combo, macro)

# Attach site location specific info
locv <- c("CropSampleID", "SoilSampleID", "lat", "lon", "crop", "fert") 
locs <- codat[locv]
site <- cbind(locs, combo)

# Write file
write.csv(site, "./MW_coda/Results/MW_site_coda.csv", row.names = F)
```

# Logratio and endmember data selections


