---
title: CoDA workflows for modeling mineral nutrient compositions of cereal grains
author: M.G. Walsh, R. Rodríguez Iglesias, R. Manners, P.C. Nalivata and M.R. Broadley
date: "Last compiled on `r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 1
    fig_caption: true
    css: style.css
---

```{r, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction

Most models of biogeography, species distributions, photosynthesis, nutrient uptake, yield, trophic dynamics, crop management and/or small area dietary requirements assume no interactions between the essential mineral nutrient elements and typically only describe the associated processes as a function of the potentially most limiting parts (i.e., the [Blackman limitation](https://www.biologydiscussion.com/photosynthesis/blackmans-law-of-limiting-factors-and-his-criticism-photosynthesis/23006)). However, the Blackman limitation does not appear to hold generally. In agricultural systems, the overall (multivariate) mineral nutrient composition is likely to affect both crop yields as well as the overall nutritional quality of plant products that are subsequently consumed by animals, including humans. The objective of this notebook is to provide starter code in [R](https://www.r-project.org/) to explore and report the mineral nutrient compositions of the main cereal grains (maize, millet, rice and sorghum) that are grown in Malawi ([Gashu et al., 2021](https://doi.org/10.1038/s41586-021-03559-3)). 

*Compositional Data Analysis* (CoDA) refers to the analyses of *compositional data* (CoDa). CoDa are defined as vectors with strictly positive components whose sum is constant (e.g, 1 or 100%). The terms also cover all other *parts* of a whole (so-called *subcompositions*), which carry only relative information ([van den Boogaart and Tolosana-Delgado, 2013](https://link.springer.com/book/10.1007/978-3-642-36809-7)). Because a change in any proportion of a whole changes at least one other proportion, CoDa are interdependent. Arguably, all concentration data should be treated CoDa because as a whole they must sum to their respective per unit totals of all constituent parts e.g., in the case of mg kg^-1^ data, to 10^6^ ppm. Therefore, CoDa are intrinsically multivariate, and their individual components should probably not be analyzed and interpreted without relating them to one another.

The original approach to CoDA developed by [J. Aitchison (1982)](https://rss.onlinelibrary.wiley.com/doi/10.1111/j.2517-6161.1982.tb01195.x) uses logarithmic ratios (LRs) of the compositional parts as the fundamental starting point for CoDa description and modeling. There is also a review paper by [M. Greenacre (2021)](https://www.annualreviews.org/doi/pdf/10.1146/annurev-statistics-042720-124436) openly available, which provides excellent guidance, as well as [R scripts](https://github.com/michaelgreenacre/CODAinPractice), about the practical applications and interpretation of CoDa. This notebook is maintained on [Github](https://github.com/mgwalsh/Bioavailability/blob/master/CoDa_workflow.Rmd), and you can fork and alter it from there for your reference and as you see fit.

# General data setup

The data we will be using were generated by the [GeoNutrition project](http://www.geonutrition.com) in Malawi. They consist of 1,808 cereal grain samples that were collected for maize (1,606), millet (31), rice (54), and sorghum (117) in proportion of their national geographical occurrence. The nationally representative cropland sampling frame and the laboratory methods that were used in the data collections are described in [Gashu et al., (2021)](https://doi.org/10.1038/s41586-021-03559-3). 

The data include [ICP-MS](https://en.wikipedia.org/wiki/Inductively_coupled_plasma_mass_spectrometry) concentration measurements of the respective cereal mineral nutrient contents (in ppm), including: Na, Mg, P, S, K, Ca, Cr, Mn, Fe, Co, Cu, Zn, Se, Mo, and I. They also include the georeference of where the samples were collected (lon/lat, EPSG:3857), the dominant cereal crop at that location, and if, or not, fertilizer and/or lime were used in growing the crop. Note that any measurements below their respective detection limits (LOD) have been set to 2/3 of the respective minimum observed values. Other imputation methods are also possible in this context see [Greenacre (2021)](https://www.annualreviews.org/doi/pdf/10.1146/annurev-statistics-042720-124436).

## Data download

To actually run this notebook, you will need to load the packages indicated in the chunk directly below to download and assemble the data. The chunk (not shown but in the markdown document) also generates an overview map of where the samples were collected in Malawi ... you can click and zoom into the individual sampling locations.

```{r, results = 'hide'}
# Package names
packages <- c("osfr", "dplyr", "leaflet", "compositions", "table1", "MASS", "archetypes")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))

# Create a data folder in your current working directory
dir.create("MW_coda", showWarnings=F)
setwd("./MW_coda")
dir.create("Results")

# Data download from OSF at: https://osf.io/btxyc/
osf_retrieve_file("btxyc") %>% osf_download(conflicts = "overwrite")
codat <- read.table("MW_grain_coda.csv", header=T, sep=",")
```

```{r, echo = FALSE}
# Grain sample locations
w <- leaflet() %>%
  setView(lng = mean(codat$lon), lat = mean(codat$lat), zoom = 7) %>%
  addProviderTiles(providers$OpenStreetMap.Mapnik) %>%
  addCircleMarkers(codat$lon, codat$lat, clusterOptions = markerClusterOptions())
w ## plot widget 
```

## CoDa transforms

This section of the notebook sets up a CoDa data frame that includes 2 amalgamations: mineral **macro-nutrients** (i.e., Na, Mg, P, S, K, and Ca ... in atomic number order), and **micro-nutrients** (i.e., Cr, Mn, Fe, Co, Cu, Zn, Se, Mo, and I). Note that the distinction between *macro* and *micro* here is only intended to reflect the recommended daily dietary values (DDV) that should be consumed by people (with the exception of Co, which should only be consumed as Vit. B12), and not their overall nutritional importance and/or benefits. The compositional macro / micro differences may also prove to be useful for measurement outlier screening, spatial mapping and/or soil property and MIR spectral predictions of nutrient compositions at a later stage. Those subsequent steps will be described in separate notebooks.

This next chunk [closes](https://econ-papers.upf.edu/papers/1554.pdf) the CoDa, and calculates *centered logratios* ([CLR](https://www.rdocumentation.org/packages/compositions/versions/2.0-1/topics/clr)). It also generates a summary table and graphs for the CLR transformed data. These can then be analyzed, summarized and interpreted with conventional univariate, multivariate, and data modeling methods ([Greenacre, 2021](https://www.annualreviews.org/doi/pdf/10.1146/annurev-statistics-042720-124436)). Note that you could also use other logratios such as the [ALR](https://www.rdocumentation.org/packages/compositions/versions/2.0-1/topics/alr), or the [ILR](https://www.rdocumentation.org/packages/compositions/versions/2.0-1/topics/ilr) and/or different amalgamations of subcompositions, but we shall leave those possibilities for you to explore. We also write out the resulting data frame to your `./MW_coda/Results` directory should you wish to process it in software other than R.

```{r}
# Calculate the CLRs
varc <- c("Na", "Mg", "P", "S", "K", "Ca", "Cr", "Mn", "Fe", "Co", "Cu", "Zn", "Se", "Mo", "I")
combo <- codat[varc]
combo <- combo / rowSums(combo) ## close the composition
combo <- as.data.frame(clr(combo)) ## centered log ratio (clr) transform

# Define the (macro / micro) amalgamation and add it to the overall nutrient composition
varm <- c("Na", "Mg", "P", "S", "K", "Ca")
macro <- rowSums(combo[varm])
combo <- cbind(combo, macro)

# Attach site data info
locv <- c("CropSampleID", "SoilSampleID", "lat", "lon", "crop", "fert") 
locs <- codat[locv]
site <- cbind(locs, combo)

# Write file
write.csv(site, "./MW_coda/Results/MW_site_coda.csv", row.names = F)

# Summary table
table1(~ Na+Mg+P+S+K+Ca+Cr+Mn+Fe+Co+Cu+Zn+Se+Mo+I+macro | crop, 
       caption = "Summary table of the CLR transformed data by crop species.", 
       overall = FALSE, data = site)
```
\
You can also plot any CLR differences between crop species with something like this next chunk.

```{r, fig.align = "center", figures-side, fig.show = 'hold', fig.fullwidth = TRUE, fig.cap = "Boxplots of micro-nutrient CLR differences between crop species."}

par(mfrow = c(3,3), pty="s", mar=c(1.8,1.8,1,1.5))
boxplot(site$Cr ~ site$crop, notch = T, ylab = "Cr", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
boxplot(site$Mn ~ site$crop, notch = T, ylab = "Mn", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
boxplot(site$Fe ~ site$crop, notch = T, ylab = "Fe", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
boxplot(site$Co ~ site$crop, notch = T, ylab = "Co", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
boxplot(site$Cu ~ site$crop, notch = T, ylab = "Cu", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
boxplot(site$Zn ~ site$crop, notch = T, ylab = "Zn", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
boxplot(site$Se ~ site$crop, notch = T, ylab = "Se", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
boxplot(site$Mo ~ site$crop, notch = T, ylab = "Mo", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
boxplot(site$I ~ site$crop, notch = T, ylab = "I", xlab = "", cex.axis = 0.7, cex.lab = 1.2)
```

# Compositional differences between crop species

This section provides initial data model summaries of parts of the CLR transformed data. We focus on the differences of the macro / micro nutrient subcompositions between crops. This can be readily extended to individual nutrients and/or other compositional amalgamations. This next chunk fits the differences in the `macro` nutrient CLRs between the 4 crop species.

```{r}
# Macro-nutrients by crop species
m0 <- glm(macro~crop, data = site)
summary(m0)
```

Based on this, there appear to be some clear differences between the macro / micro compositional parts of the 4 cereal crops. Maize appears to have a significantly higher macro-nutrient proportion than millet, rice or sorghum. Conversely, because of the binary macro / micro contrast imposed here, the proportion of total micro-minerals would be expected to be proportionally higher in millet, rice and sorghum as compared to maize. The next chunk includes the use of fertilizer and/or lime.

```{r}
# Macro-nutrients by crop species and fertilizer / lime use
m1 <- glm(macro~crop+fert, data = site)
summary(m1)
```

Note that the interactions between crop species and fertilizer / lime applications are difficult to estimate based on the available Malawi dataset that is dominated by maize. An oversample of the more locally common millet, rice and sorghum grown with regard to fertilizer use might be useful in this context. In the next chunk we fit a main effects model that looks at the zinc (`Zn`) CLR part of the micro-nutrient subcomposition. Note that we include the `macro` subcomposition CLR as a covariate, to retain a link to the overall 15 element composition. Similar analyses could be done easily for the remaining parts.

```{r}
# Zn part by crop species and fertilizer / lime use
Zn <- glm(Zn~macro+crop+fert, data = site)
summary(Zn)
```

Note that the inclusion of the `macro` subcomposition here takes care of a lot of the between crop species variation in the Zn CLR values. Similar analyses could easily be done for the remaining nutrient parts.

# Compositional archetypes

*Archetypal analysis* ([Cutler and Breiman, 1994](https://www.stat.cmu.edu/technometrics/90-00/vol-36-04/v3604338.pdf)) is an unsupervised learning method similar to cluster analysis. It describes each observation in a dataset set as a mixture of archetypes. Rather than describing the average or *typical* observations and measurements, archetypal analysis seeks the extreme points in multidimensional data, which can provide a simple and useful way of looking at multivariate data. The identified archetypes themselves are generally mixtures of the individual variables under consideration. Archetypal mixture coefficients and projections are also CoDa, which is why we include them in this notebook. We will consider them as subcompositions.

# Takeways

The main takeaways from this notebook are the following:

* Converting nutrient concentrations to logratios is a simple step, but unlike other data transformations it conserves the integrity of the relevant compositional parts needed to normalize the data for any subsequent univariate, multivariate and/or inferential analyses and statistics. Also, the CoDa can always be back-transformed into their original units.

* We think that it may be more informative to analyze mineral nutrient concentration data as CoDa. The associated LR conversions (ALR, CLR, ILR) do not make the data any more computationally difficult, but they should improve any data modeling based inferences where subcompositional closure principles apply.

Any questions or comments about this notebook are most welcome via [AFSIS](mailto:mgw.africasoils.info).
