---
title: Spectrometry workflows for predicting Zinc levels in Maize grain and soils
author: M.G. Walsh, P.C. Nalivata, E. Towett and M. Broadley
date: "Last compiled on `r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 1
    fig_caption: true
    css: style.css
---

```{r, echo=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# Introduction

Micronutrients i.e., the essential vitamins and minerals, are vital for healthy development and disease prevention in people. With the exception of Vitamin D, micronutrients are not produced in the body and must be obtained from the diet. Though people only need small amounts of micronutrients, consuming their recommended amounts is important, as *micronutrient deficiencies* (MNDs) can have severe consequences. However, national surveys containing dietary intake data are not routinely collected in most developing countries because they are expensive and time consuming to obtain both in the field and the laboratory. Furthermore, the databases that are needed to analyze dietary intakes should be geographically explicit to be effective but they are often lacking and/or not up-to-date. Similarly, agronomic interventions that focus on plant micronutrients for improving cropland productivity are inadequately supported by spatially/temporally explicit data.

Zinc deficiency is among the 5 most widespread MNDs (including Iron, Vitamin A, Iodine, Folate and *Zinc*) of global concern [Bailey et al (2015)](https://doi.org/10.1159/000371618), and is the focus of this notebook. Zinc (Zn) is a trace element found in varying concentrations in all soils, plants and animals. It is essential for the normal growth of higher plants, animals and humans. While Zn is needed in only small quantities, if those amounts are inadequate, plants and/or animals will suffer from physiological stress brought about by the dysfunction of several enzyme systems and other metabolic functions in which Zn plays a key role [Wessels and Brown (2012)](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0050568). In Malawi, Zn deficiency prevalence rates of ~60% in people have been reported based on a national survey of serum Zn concentrations ([Likoswe et al., 2020](https://doi.org/10.3390/nu12061563)). This value may be even higher in rural areas ([Siyame et al., 2013](https://doi.org/10.1024/0300-9831/a000158)).

The spectral signatures of soils, plants and materials generally are defined by their reflectance or absorbance as a function of wavelength in the electromagnetic spectrum. Under controlled conditions, the signatures result from electronic transitions of atoms and vibrational stretching and bending of structural groups of atoms that form molecules or crystals. Spectroscopy has been shown to provide highly repeatable, rapid and low cost measurements of many different soil and plant properties in numerous studies. The amount of light absorbed by a sample can be measured with minimal sample preparation (primarily drying and fine grinding) across a wide range of ultra-violet (UV), visible (VIS), near (NIR) and mid-infrared (MIR) wavelengths to provide a unique spectral signature. An individual measurement can be performed in about 30 seconds, in contrast to conventional soil and plant tests, which are typically slow, labor-intensive, expensive and/or use hazardous chemicals. 

The main empirical questions posed in this notebook are twofold:

* Can Zn concentrations in soils be used to predict Zn concentrations in Maize grain? While there must be kinetic links between soils and plants, the uptake and bioavailability of Zn for food crops are generally complex and not well quantified or predicted. 

* Can we use MIR spectra to predict Zn concentrations in Maize grain and soils reliably? If so, this would advance and enable localized Zn deficiency diagnostics and risk assessments that could potentially be carried out routinely, at low cost and across large geographical regions and populations of interest. 

The notebook is intended for self-learning in [R](https://www.r-project.org/). It does not go into the details of the [spectroscopy](https://en.wikipedia.org/wiki/Spectroscopy) itself (which is well covered in the literature), but instead it focuses on [spectrometry](https://en.wikipedia.org/wiki/Spectrophotometry) processes and the associated computing workflows that are needed to generate useful predictions from a population of spectral signatures of  (features) relative to their corresponding reference measurements (labels). In this particular example, we shall use maize grain Zn concentration data and co-located topsoil (0-20 cm) data that were collected as part of the [GeoNutrition project](http://www.geonutrition.com) in Malawi.

# General data setup

To actually run this notebook, you will need to load the packages indicated in the chunk directly below. This allows you to assemble the wet chemistry and spectral dataframes providing a lot of options to generate spectral predictions of Zn plant and soil micro-nutrient. The notebook itself is maintained on [Github](https://github.com/mgwalsh/Bioavailability/blob/master/Zn_preds.Rmd), and you can fork and modify it from there as you see fit.

```{r}
# Package names
packages <- c("downloader", "caret", "caretEnsemble", "soil.spec", "MASS", "pls", "glmnet", "randomForest", "xgboost", "Cubist", "quantreg", "leaflet", "htmlwidgets", "plyr", "dplyr", "doParallel")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

The next chunk downloads the data needed for running this example. It assembles georeferenced soil measurements collected across Africa and links these to [FT-IR](https://www.bruker.com/en/products-and-solutions/infrared-and-raman/ft-ir-routine-spectrometer/alpha-ii-compact-ft-ir-spectrometer.html) spectra. Note that this chunk is Mac or Linux specific and so the directory structures would need to be changed slightly to run on Windows machines.

```{r}
# Create a data folder in your current working directory
dir.create("MW_Zn", showWarnings = F)
setwd("./MW_Zn")
dir.create("Results", showWarnings = F)
```

The following chunk then writes out the dataframe `MW_Zn_data.csv` into your `./MW_Zn/Results` directory if you'd like to process those outputs in software other than R. It also generates a location map of where those soil samples were obtained.

